%!TEX root = ../dissertation.tex
% the acknowledgments section

\newthought{I wish} to thank my dissertation advisor, Professor John Quackenbush, whose leadership, support, wisdom and perspective sets the gold standard for any student fortunate enough to mentored by him. He was always available to patiently guide me through my research and continuously replenished my passion for science along the way with his own excitement. He inspired me in the fight against cancer (literally and figuratively) and served as a role model for perseverance, ambition and attitude in the long road to the PhD.

I am grateful to my committee members, Kimberly Glass and Christoph Lange, whose patience and expertise were invaluable in my progression through the program. Both spent countless hours with me discussing, planning, guiding and editing. These chapters could not have been written without them.

I am incredibly grateful for the network of students, faculty and staff in the Biostatistics department who all contributed positively to the creation of a community that fosters growth and challenges us all to succeed. Additionally, I want to thank the postdocs and grad students of the Quackenbush lab – Joe Paulson, John Platig, Marieke Kuijjer, Joe Barry, Maude Fagny, Camila Lopes-Ramos, Megha Padi, Joey Chen, and Heather Selby for making the group a warm, welcoming, helpful, exciting, and creative environment that I looked forward to working in every day.

I want to thank my amazing friends and family, in particular my Dad, Mike and Amy for their love and support through these last five years. Jen, who stood with me throughout the years and is my motivation to persevere. Also, my dog Tails and my cat Willow, who provided critical emotional support. And finally, Mom, who reminded me and continues to be a reminder everyday to strive to be the best version of myself that I can be.